Server-Side Probabilistic Programming

### 2018-11-01 abstract/paper.tex
cosmetic

\begin{abstract}
	We share our experiences from integrating probabilistic
	programming into a server-side software system and
	present our ongoing work on probabilistic programming
	facility for Go, a modern programming language of choice for
	server-side software development. We demonstrate how a
	lightweight efficient probabilistic programming facility can
	be added to an existing programming language.  Server-side
	application of probabilistic programming poses challenges
	for a probabilistic programming system. We discuss the
	challenges and our experience in overcoming them, and
	suggest guidelines which can help in a wider adoption of
	probabilistic programming in server-side software systems.
\end{abstract}

@@

\paragraph{Simulation-inference dichotomy} Probabilistic models
often follow the design pattern of simulation-inference: a
significant part of the model is a simulator, running an
algorithm with fixed parameters; the optimal parameters, or
their distribution, are to be inferred. The inferred parameters
are then used by the software system to execute the simulation
independently of inference for forecasting and decision making.

This pattern suggest re-use of the simulator: instead of
implementing the simulator twice, in the probabilistic model and
in the host system, the same code can serve both purposes.
However to achieve this, the host language must coincide with
the implementation language of the probabilistic model, on one
hand, and allow a computationally efficient implementation of
the simulation, on the other hand. Some probabilistic systems
(Figaro~\cite{P09}, Anglican~\cite{TMY+16}, Turing~\cite{GXG18})
are built with tight integration with the host environment in
mind; more often than not though the probabilistic code is
not trivial to re-use.

@@

\paragraph{Integration and deployment} Deployment of server-side
software systems is a delicate process involving automatic
builds and maintenance of dependencies. Adding  a component
which possibly introduces additional software dependencies or
even a separate runtime complicates deployment. Minimizing the
burden of probabilistic programming on the integration
and deployment process should be a major consideration in
design or selection of probabilistic programming tools.
Probabilistic programming systems which are implemented or 
provide an interface in a popular programming language, e.g.
Python (Edward~\cite{THS+17}, Pyro~\cite{Pyro18}) are easier
to integrate and deploy, however the smaller the footprint
of a probabilistic system, the easier is the adoption.

@@

\begin{enumerate}
	\item The Go parser and abstract syntax tree serializer are a
		part of the standard library. Reading, transforming, and
		writing Go source code is straightforward and effortless.
	\item Type inference (or \textit{type checking} as it is
		called in the Go ecosystem), also provided in the
		standard library, augments parsing and allows to
		selectively apply transformation-based automatic
		differentiation  based on static expression types. 
	\item Go compiles and runs fast. Fast compilation and
		execution speeds allow to use the same facility for both
		exploratory design of probabilistic models and for
		inference in production environment.
	\item Go offers efficient parallel execution as a
		first-class feature, via so-called \textit{goroutines}.
		Goroutines streamline implementation of sampling-based
		inference algorithms. Sample generators and consumers
		are run in parallel, communicating through channels. 
		Inference is easy to parallelize in order to exploit
		hardware multi-processing, and samples are retrieved
		lazily for postprocessing. 
\end{enumerate}

>>>>

\begin{abstract}
	We share our experiences from integrating probabilistic
	programming into a server-side software system and
	present our ongoing work on probabilistic programming
	facility for Go, a modern programming language of choice for
	server-side software development. We demonstrate how a
	lightweight efficient probabilistic programming facility can
	be added to an existing programming language.  Server-side
	application of probabilistic programming poses challenges
	for a probabilistic programming system. We discuss the
	challenges and our experience in overcoming them, and
	suggest guidelines that can help in a wider adoption of
	probabilistic programming in server-side software systems.
\end{abstract}

@@

\paragraph{Simulation-inference dichotomy} Probabilistic models
often follow a design pattern of simulation-inference: a
significant part of the model is a simulator, running an
algorithm with fixed parameters; the optimal parameters, or
their distribution, are to be inferred. The inferred parameters
are then used by the software system to execute the simulation
independently of inference for forecasting and decision making.

This pattern suggest re-use of the simulator: instead of
implementing the simulator twice, in the probabilistic model and
in the host environment, the same code can serve both purposes.
However to achieve this, the host language must coincide with
the implementation language of the probabilistic model, on one
hand, and allow a computationally efficient implementation of
the simulation, on the other hand. Some probabilistic systems
(Figaro~\cite{P09}, Anglican~\cite{TMY+16}, Turing~\cite{GXG18})
are built with tight integration with the host environment in
mind; more often than not though the probabilistic code is
not trivial to re-use.

@@

\paragraph{Integration and deployment} Deployment of server-side
software systems is a delicate process involving automatic
builds and maintenance of dependencies. Adding  a component,
which possibly introduces additional software dependencies or
even a separate runtime, complicates deployment. Minimizing the
burden of probabilistic programming on the integration
and deployment process should be a major consideration in
design or selection of probabilistic programming tools.
Probabilistic programming systems that are implemented or 
provide an interface in a popular programming language, e.g.
Python (Edward~\cite{THS+17}, Pyro~\cite{Pyro18}) are easier
to integrate and deploy, however the smaller the footprint
of a probabilistic system, the easier is the adoption.

@@

\begin{itemize}
	\item The Go parser and abstract syntax tree serializer are a
		part of the standard library. Reading, transforming, and
		writing Go source code is straightforward and effortless.
	\item Type inference (or \textit{type checking} as it is
		called in the Go ecosystem), also provided in the
		standard library, augments parsing and allows to
		selectively apply transformation-based automatic
		differentiation  based on static expression types. 
	\item Go compiles and runs fast. Fast compilation and
		execution speeds allow to use the same facility for both
		exploratory design of probabilistic models and for
		inference in production environment.
	\item Go offers efficient parallel execution as a
		first-class feature, via so-called \textit{goroutines}.
		Goroutines streamline implementation of sampling-based
		inference algorithms. Sample generators and consumers
		are run in parallel, communicating through channels. 
		Inference is easy to parallelize in order to exploit
		hardware multi-processing, and samples are retrieved
		lazily for postprocessing. 
\end{itemize}
